---
title: "Basic model of evolutionary rescue in discrete time"
author: "Claudia Bank"
date: "2025-10-14"
output:
  html_document:
    df_print: paged
---

------------------------------------------------------------------------

## Simulating evolutionary rescue in discrete time

In this notebook, we are implementing simulations of evolutionary rescue by a single mutation in discrete time, as discussed and analyzed in the work by Orr & Unckless. The model is as follows: a haploid population starts with a given number of wild-type and mutant individuals. Because of a sudden environmental change, the wild type is unfit and declines in the new environment. The mutant is beneficial and can survive and expand in the new environment, but it starts at a low (or zero) frequency in the total population. In order to rescue the population, it has to outcompete the wild type. If the mutant is not present at time 0, it has to appear through mutation and then spread in the population before extinction occurs.

## List of variables and parameters

-   **N_end_wd:** size of the mutant (wd) population

-   **N_end_wb:** size of the mutant (wb) population

-   **N_end_sd:** size of the mutant (sd) population

-   **N_end_sb:** size of the mutant (sb) population

-   **init_wd:** initial population size of population with the weak deleterious mutation (should be a positive integer)

-   **init_wb:** initial population size of the population with the weak beneficial mutation (should be a non-negative

-   **init_sd:** initial population size of the population with the strong deleterious mutation (should be a positive integer)

-   **init_sb:** initial population size of the population with the strong beneficial mutation (should be a non-negative integer)

-   **decay_rate_wm:** detrimental effect of the environment on the weak mutation (i.e., the populations carrying weak mutations decays at this rate; should be \>0 to induce potential extinction)

-   **decay_rate_sm:** detrimental effect of the environment on the strong mutation (i.e., the populations carrying strong mutations decays at this rate; should be \>0 to induce potential extinction)

-   **s_wm:** selective strength of the weak mutation (should be \>decay_rate to allow for evolutionary rescue)

-   **s_sm:** selective strength of the strong mutation (should be \>decay_rate to allow for evolutionary rescue)

-   **m_rate_wm:** mutation rate from the weak deletrious to the weak beneficial deleterious mutation and from weak beneficial to weak deleterious mutation (should be \>0 and small)

    **m_rate_sm:** mutation rate from the strong deleterious to the strong beneficial mutation and from strong beneficial to strong deleterious mutation (should be \>0 and small)

-   **max_gen:** number of generations after which to stop the simulation (should be an integer \>0; this is not a model parameter but a setting for the simulations). In the code presented here, the simulation stops automatically if the population goes extinct, or if it recovers and exceeds 10 times the original total population size, so **max_gen** can be set to a large value without creating extremely long simulation times.

## Simulate one generation of population size change

We want to observe the numbers of wild-type and mutant individuals over time. As first step, we simulate one generation of the population. At time $t$, the population has $N_a$ wild-type individuals and $N_A$ mutant individuals. First, these reproduce. This is modeled by drawing a Poisson random variable according to the expected absolute number $N_a \cdot (1-r)$ of wild-type individuals, where $r$ is the decay rate of the wild type, and $N_A \cdot (1-r+s)$ of mutant individuals, where $s$ is the selective advantage of the mutant over the wild time. Then mutation from $a$ to $A$ occurs as a Poisson random variable according to the given mutation rate. This defines the composition of the population at time $t+1$.

## Our Question

How do 2 deleterious mutations of different strength , which can mutate to beneficial ones, influence the probability of rescue and how does their frequency in the population change?
How do 2 mutations, which can alternate between beneficial and deleterious versions of the same strength, influence the probability of rescue and how does their frequency in the population change?
## Definitions

-   wd: weak deleterious
-   wb: weak beneficial
-   sd: strong deleterious
-   sb: strong beneficial

```{r}
# Function for simulating one Generation in the Population
# adding the population sizes, decay rates, seleection coefficients and mutation rates
simulate_one_gen <- function(N_wd, N_wb, N_sd, N_sb, decay_rate_wm, decay_rate_sm, s_wm, s_sm, m_rate_wm, m_rate_sm) {
  
  # draw Offspring according to Poisson Distribution
  # altered so all four different mutations are considered 
  # for deleterious mutations the s is calculated minus (negative effect) and for the beneficial muation the s is added (positive effect)
  
  offsp_wd <- rpois(1, N_wd * (1-decay_rate_wm-s_wm))
  offsp_wb <- rpois(1, N_wb * (1-decay_rate_wm+s_wm))
  offsp_sd <- rpois(1, N_sd * (1-decay_rate_sm-s_sm))
  offsp_sb <- rpois(1, N_sb * (1-decay_rate_sm+s_sm))
  
  # draw changing Mutants according to Poisson Distribution
  # alterations considering the for different mutations 
  
  mut_wd_to_wb <- rpois(1, offsp_wd * m_rate_wm)
  mut_sd_to_sb <- rpois(1, offsp_sd * m_rate_sm)
  mut_wb_to_wd <- rpois(1, offsp_wb * m_rate_wm)
  mut_sb_to_sd <- rpois(1, offsp_sb * m_rate_sm)
  
  # determine new Population Sizes of all four Mutants (2 versions of weak Mutations   and 2   versions of strong Mutations)
  # to account for back mutations as well (beneficial to deleterious), mutations rates for these cases are added (mut_b_to_d)
  
  N_wd_new <- max(offsp_wd - mut_wd_to_wb+ mut_wb_to_wd,0)
  N_wb_new <- max(offsp_wb - mut_wb_to_wd+ mut_wd_to_wb,0)
  N_sd_new <- max(offsp_sd - mut_sd_to_sb+ mut_sb_to_sd,0)
  N_sb_new <- max(offsp_sb - mut_sb_to_sd+ mut_sd_to_sb,0)
  
  # with the return function, the new values for population sizes are stored
  
 return(c(N_wd_new, N_wb_new, N_sd_new, N_sb_new))
}
  # with print we visualize the 4 numbers we get for the population sizes, we simulate the population by setting our parameters to certain values 


print(simulate_one_gen(N_wd=50, N_wb=0, N_sd=50, N_sb=0, decay_rate_wm=0, decay_rate_sm=0, s_wm=0, s_sm=0, m_rate_wm=0, m_rate_sm=0))
```

## Simulate a population trajectory

We can now track the dynamics of the population from one generation to another, but this is not enough. We want to follow the population from time 0 until it either is rescued or goes extinct. This is what the following function does.

```{r}
# Simulation runs until the 100th Generation

max_gen=1000

#With this function we simulate over generations, we put initial population sizes of the four mutations, with the parameter max_gen we determine that our simulation maximally runs until the 100th geneeration

simulate_pop <- function(init_wd, init_wb,  init_sd, init_sb, decay_rate_wm, decay_rate_sm, s_wm, s_sm, m_rate_wm, m_rate_sm, max_gen) {
  
  # Create the Vector in which to save the Results
  
  pop_vector <- c(init_wd,init_wb, init_sd, init_sb)
  
  # initiate the variables
  
  pop_new <- c(init_wd, init_wb, init_sd, init_sb)
  
  # run the Simulation until Generation max_gen
  
  for (i in 1:(max_gen + 1)) {
    
    # redefine the current Population one Generation later
    
    pop_new <- simulate_one_gen(pop_new[1],pop_new[2],pop_new[3],pop_new[4], decay_rate_wm, decay_rate_sm, s_wm, s_sm, m_rate_wm, m_rate_sm)
    
    # add the new Population Sizes to the Output Vector
    
    pop_vector <- rbind(pop_vector,pop_new)
    
    # Condition to stop the Simulation before max_gen: either the Population exceeds 10 Times the Original Population Size, or it goes extinct
    
    if ((sum(pop_new) >= 10 * sum(c(init_wd, init_wb, init_sd, init_sb))) || sum(pop_new) == 0)
      break

}
  
  # Define the Row and Column Names of the Output Vector
  rownames(pop_vector) <- (0:max_gen)[1:length(pop_vector[,1])] 
  
  # the colom-names in our vectors are according to our mutations (wd, wb, sd, sb)
  
  colnames(pop_vector) <- c("wd","wb", "sd", "sb")
  
  # return the vector
  
  return(pop_vector)	
}


# Test the Function and plot the Result

max_gen <- 100

# Simulation Data with certain values determined

output <- simulate_pop(init_wd=30, init_wb=30,  init_sd=30, init_sb=30, decay_rate_wm=0.1, decay_rate_sm=0.2, s_wm=0.1, s_sm=0.2, m_rate_wm=0.003, m_rate_sm=0.005,max_gen)

# show the last few Lines of the Data Table

print(tail(output))

# plot the Output 
# determine x axis Range 

x_range <- 0:(length(output[,1])-1)

# this plots the total Population Size-black line visible in the plot

plot(x_range,output[,1]+output[,2]+output[ ,3]+output[ ,4],type='l',ylim=c(0,max(output[,1]+output[,2]+output[ ,3]+output[ ,4])),xlab =  "Generation",ylab = "Population size")

# add Number of wd Individuals
lines(x_range,output[,1], col="#77b5fe")

# add Number of wb Individuals
lines(x_range,output[,2], col="navyblue")

# add Number of sd Individuals
lines(x_range,output[,3], col="#D55E00")

# add Number of sb Individuals
lines(x_range,output[,4], col="darkred")

# add Legend in the top-left corner
legend("topleft",
       legend=c("Weak deleterious","Weak beneficial","Strong deleterious","Strong beneficial"),
       col=c("#77b5fe","navyblue","#D55E00","darkred"), lty=1, lwd=1.5)





```
```{r}
# code for plotting the result of different populations in one plot
init_wd <- 50
init_wb <- 50
init_sd <- 50
init_sb <- 50
m_rate_wm <- 0.002 # approximately 0.4
m_rate_sm <- 0.003 # approximately 0.6
decay_rate_wm <- 0.1
decay_rate_sm <- 0.2
s_wm <- 0.1
s_sm <- 0.2
max_gen <- 1000
replicates <- 30


# defining Data Frame to save one Row (Population Size for multiple Generations until max_gen) after one Loop
temp_output <- data.frame(array(NA, dim = c(0,max_gen)))
# defining Data Frame to save all Results from Loop
final_output <- data.frame(array(NA, dim = c(replicates,max_gen)))
# defining a Set of colors in order to color each Line differently
lines_colors <- rainbow(replicates)

# simulating a population multiple times
for (i in 1:replicates) {
  temp_output <- rowSums(simulate_pop(
    init_wd, init_wb, init_sd, init_sb,
    decay_rate_wm, decay_rate_sm,
    s_wm, s_sm,
    m_rate_wm, m_rate_sm,
    max_gen = max_gen)[, c("wd", "wb", "sd", "sb")])
  
  final_output[i, 1:length(temp_output)]<-temp_output
}

# defining end_gen which is as long as the last Generation where the pop size is an actual Number and not NA yet
end_gen<-tail(which(!is.na(final_output[1,])), 1)
# Plot one Line / one Population
plot(x=1:max_gen,y=final_output[1,1:max_gen],type='l',ylim=c(0,max(final_output, na.rm = TRUE)),xlab =  "Generation",ylab = "Population size", col = lines_colors[1])
# Loop for plotting each Row of Data Table / Population as a Line in the Plot
for (i in 2:replicates) {
  
  end_gen<-tail(which(!is.na(final_output[i,])), 1)
  lines(x=1:max_gen,y=final_output[i,1:max_gen], col=lines_colors[i])
  
  
}


```
## A small analysis

As you can see if you simulate and plot the output several times, there is a lot of stochasticity/variation in the outcome. Therefore, to get interpretable results, we have to run the simulation many times and summarize the outcome depending on the question of our study. Here I am showing an example in which we record the time and the value of the lowest population size (or the time of extinction, if the population size eventually goes to 0). Usually, we want to study this for a range of model parameters; here I am varying the decay rate and the selection coefficient. Other (maybe more interesting) features to study would be the following: =\>simulating varying parameters with replicates per one value

-   rescue probability dependent on initial frequency and selection coefficient
-   time at which the rescue mutation occurs first if it rescues, vs. if it doesn't rescue, dependent on initial population size and mutation rate
-   time until the population has recovered in case it becomes rescued, dependent on mutation rate and selection coefficient
-   ...

```{r}
# set some Parameters to fixed Values
init_wd <- 30
init_wb <- 25
init_sd <- 30
init_sb <- 25
m_rate_wm <- 0.01
m_rate_sm <- 0.05
decay_rate_wm <- 0.1
decay_rate_sm <- 0.2
s_wm <- 0.2
s_sm <- 0.4
max_gen <- 1000
# determine how often to run the Simulation for each set of Parameters
no_replicates <- 10


# set Parameters to vary
s_values_wm <- c(0.10,0.15,0.2)
s_values_sm <- c(0.2,0.205,0.210)
#m_values_wm <- c(0.001,0.01,0.05)
#m_values_sm <- c(0.05,0.1,0.15)


# initialize Data Table - where to collect the Results
data_table <- c()
data_table_wd<- c()
data_table_wb<- c()
data_table_sd<- c()
data_table_sb<- c()
# empty list to save rescue results
rescue_results <- data.frame(
  s_wm = numeric(),
  s_sm = numeric(),
  replicate = integer(),
  rescued = integer()
)
# define total runs and rescue count
total_runs<-0
rescue_count<-0
# run the Simulation across all chosen parameters
# loop over mutation rates
#for(mval_wm in m_values_wm){
 # for(mval_sm in m_values_sm){
    # loop over Selection strength
    for(sval_wm in s_values_wm){
      for(sval_sm in s_values_sm){
        # different Way of running many Simulations: make Replicates using "repeat" Function         with a counter i
        # reset Counter
        i<-1
        
        repeat {
          # increase Counter by one
          i<-i+1
          # defining total runs
          total_runs <- total_runs + 1
          # run the Simulation once
          one_run <- simulate_pop(init_wd, init_wb, init_sd, init_sb,
                                  decay_rate_wm, decay_rate_sm,
                                  sval_wm, sval_sm, m_rate_wm=0.01,m_rate_sm=0.01, max_gen)
          
          # determine total Population Sizes
          total_size <- one_run[,1]+one_run[,2]+one_run[,3]+one_run[,4]
          
          
          # determine final Population Size for each Mutant
          N_end_wd <- one_run[nrow(one_run), "wd"]
          N_end_wb <- one_run[nrow(one_run), "wb"]
          N_end_sd <- one_run[nrow(one_run), "sd"]
          N_end_sb <- one_run[nrow(one_run), "sb"]
          
          total_end <- N_end_wd + N_end_wb + N_end_sd + N_end_sb
          # introduce rescue count
          rescued_wd <- ifelse(one_run[nrow(one_run), "wd"] > 0, 1, 0)
          rescued_wb <- ifelse(one_run[nrow(one_run), "wb"] > 0, 1, 0)
          rescued_sd <- ifelse(one_run[nrow(one_run), "sd"] > 0, 1, 0)
          rescued_sb <- ifelse(one_run[nrow(one_run), "sb"] > 0, 1, 0)

         
          # save rescue outcome (rescue or extinction)
          rescue_results <- rbind(rescue_results,
            data.frame(
              s_wm = sval_wm,
              s_sm = sval_sm,
              replicate = i - 1,
              rescued_wd = rescued_wd,
              rescued_wb = rescued_wb,
              rescued_sd = rescued_sd,
              rescued_sb = rescued_sb
  )
)

          # determine Frequencies of each mutant in the final Population
          freq_wd <- ifelse(total_end > 0, N_end_wd / total_end, 0)
          freq_wb <- ifelse(total_end > 0, N_end_wb / total_end, 0)
          freq_sd <- ifelse(total_end > 0, N_end_sd / total_end, 0)
          freq_sb <- ifelse(total_end > 0, N_end_sb / total_end, 0)
          
          
          
          data_table_wd<- rbind(data_table_wd, c(sval_wm,freq_wd))
          data_table_wb<- rbind(data_table_wb, c(sval_wm, freq_wb))
          data_table_sd<- rbind(data_table_sd,c(sval_sm, freq_sd))
          data_table_sb<-rbind(data_table_sb, c (sval_sm,freq_sb))
          # enter the D into the Table
          data_table <- rbind(data_table,c(m_rate_wm = 0.01,m_rate_sm = 0.01,sval_wm,sval_sm,N_end_wd,        N_end_wb, N_end_sd, N_end_sb,freq_wd, freq_wb, freq_sd,freq_sb)) # note that we add the                varying parameters (mutation rate and selection strength for each version of weak            and strong mutation) to the table too
          # stop the repeated Computation after no_replicates times
          
          
          if(i>no_replicates) break
          
        }
      }
    }
  #}
#}
# Make sure your data is a proper data frame
data_table_wd <- as.data.frame(data_table_wd)

# If the column names are missing or wrong, set them
colnames(data_table_wd) <- c("V1", "V2")

# Add an index column for plotting along x-axis
data_table_wd$Index <- 1:nrow(data_table_wd)

# calculate rescue probability per parametercombination
library(dplyr)
library(ggplot2)
library(tidyr)



rescue_summary <- rescue_results %>%
  group_by(s_wm, s_sm) %>%
  summarise(
    rescue_prob_wd = mean(rescued_wd),
    rescue_prob_wb = mean(rescued_wb),
    rescue_prob_sd = mean(rescued_sd),
    rescue_prob_sb = mean(rescued_sb),
    .groups = "drop"
  )

print(rescue_summary)


# Daten vorbereiten
rescue_long <- rescue_summary %>%
  pivot_longer(
    cols = starts_with("rescue_prob_"),
    names_to = "mutant",
    values_to = "rescue_prob"
  ) %>%
  mutate(
    mutant = gsub("rescue_prob_", "", mutant),
    s_value = case_when(
      mutant %in% c("wd", "wb") ~ s_wm,
      mutant %in% c("sd", "sb") ~ s_sm
    )
  )

library(ggplot2)

ggplot(rescue_long, aes(x = factor(s_value), y = rescue_prob, fill = mutant)) +
  geom_boxplot(alpha = 0.7) +
  scale_y_continuous(limits = c(0, 1)) +
  scale_fill_manual(values = c(
    "wd" = "#77b5fe", 
    "wb" = "navyblue", 
    "sd" = "#D55E00", 
    "sb" = "darkred"
  )) +
  facet_wrap(~ mutant, scales = "free_x") +
  labs(
    title = "Rescue Probability per Mutant vs Relevant Selection Coefficient (Boxplot)",
    x = "Relevant selection coefficient (s)",
    y = "Rescue Probability",
    fill = "Mutant"
  ) +
  theme_minimal(base_size = 14)

ggplot(rescue_long, aes(x = s_value, y = rescue_prob, color = mutant)) +
  geom_jitter(width = 0.01, height = 0, size = 3, alpha = 0.7) +
  stat_summary(fun = mean, geom = "point", shape = 18, size = 4, color = "black") +
  scale_y_continuous(limits = c(0, 1)) +
  scale_color_manual(values = c(
    "wd" = "#77b5fe", 
    "wb" = "navyblue", 
    "sd" = "#D55E00", 
    "sb" = "darkred"
  )) +
  facet_wrap(~ mutant, scales = "free_x") +
  labs(
    title = "Rescue Probability per Mutant vs Relevant Selection Coefficient (Scatterplot)",
    x = "Relevant selection coefficient (s)",
    y = "Rescue Probability",
    color = "Mutant"
  ) +
  theme_minimal(base_size = 14)


0# optional: Mutantennamen schöner machen
rescue_long$mutant <- gsub("rescue_prob_", "", rescue_long$mutant)











# plotting wd with different s values
  # making sure that its a data frame
data_table_wd <- as.data.frame(data_table_wd)
colnames(data_table_wd) <- c("V1", "V2")


  # Scatterplot of wd with jitter and mean 
library(ggplot2)
ggplot(data_table_wd, aes(x = factor(V1), y = V2, color = factor(V1))) +
  geom_jitter(width = 0.2, height = 0, size = 2, alpha = 0.7) +
  stat_summary(fun = mean, geom = "point", shape = 18, size = 4, color = "black") +
  labs(
    x = "selection coefficients",
    y = "frequency",
    color = "selection coefficient",
    title = "Weak deleterious-Means of frequencies of the different selection coefficients"
  ) +
  theme_minimal()

# plotting wb with different s values
  # making sure that its a data frame
data_table_wb <- as.data.frame(data_table_wb)
colnames(data_table_sb) <- c("V1", "V2")

# Scatterplot of wb with jitter and mean
ggplot(data_table_wb, aes(x = factor(V1), y = V2, color = factor(V1))) +
  geom_jitter(width = 0.2, height = 0, size = 2, alpha = 0.7) +
  stat_summary(fun = mean, geom = "point", shape = 18, size = 4, color = "black") +
  labs(
    x = "selecton coefficients",
    y = "frequency",
    color = "selection coefficients",
    title = "Weak beneficial-Means of frequencies of the different selection coefficientsn"
  ) +
  theme_minimal()

# plotting sd with different s values
  # making sure that its a data frame
data_table_sd <- as.data.frame(data_table_sd)

colnames(data_table_sd) <- c("V1", "V2")

  # Scatterplot of sd with jitter and mean
ggplot(data_table_sd, aes(x = factor(V1), y = V2, color = factor(V1))) +
  geom_jitter(width = 0.2, height = 0, size = 2, alpha = 0.7) +
  stat_summary(fun = mean, geom = "point", shape = 18, size = 4, color = "black") +
  labs(
    x = "selection coefficients ",
    y = "frequency",
    color = "selection coefficients",
    title = "Strong deletrious-Means of frequencies of the different selection coefficients"
  ) +
  theme_minimal()

# plotting sb with different s values
  # making sure that its a data frame

data_table_sb <- as.data.frame(data_table_sb)
colnames(data_table_sb) <- c("V1", "V2")

  # Scatterplot of sb with jitter and mean 
ggplot(data_table_sb, aes(x = factor(V1), y = V2, color = factor(V1))) +
  geom_jitter(width = 0.2, height = 0, size = 2, alpha = 0.7) +
  stat_summary(fun = mean, geom = "point", shape = 18, size = 4, color = "black") +
  labs(
    x = "selection coefficients",
    y = "frequency",
    color = "selection coefficients",
    title = "Strong beneficial-Means of frequencies of the different selection coefficients"
  ) +
  theme_minimal()


# mean Frequencies of each mutant
mean_freq_wd <- mean(freq_wd)*100
mean_freq_wb <- mean(freq_wb)*100
mean_freq_sd <- mean(freq_sd)*100
mean_freq_sb <- mean(freq_sb)*100

print(paste("Mean frequency of wd =", mean_freq_wd) )
print(paste("Mean frequency of wb =", mean_freq_wb) )
print(paste("Mean frequency of sd =", mean_freq_sd) )
print(paste("Mean frequency of sb =", mean_freq_sb) )

# define Column Names of Data Table
colnames(data_table) <- c("m_rate_wm","m_rate_sm","s_wm","s_sm","N_end_wd", "N_end_wb", "N_end_sd", "N_end_sb","freq_wd", "freq_wb", "freq_sd","freq_sb")
# show the first and last Lines of the Output
print(head(data_table))
print(tail(data_table))

```
