---
title: "Basic model of evolutionary rescue in discrete time"
author: "Claudia Bank"
date: "2025-10-14"
output:
  html_document:
    df_print: paged
---

This notebook is meant to introduce you to simulations and calculations using R. In the following, we will simulate the evolutionary rescue model introduced in the lecture.

## Your tasks

Read through the code and execute it (using the "play" or "run" or "knit/preview" buttons - you might have to install some extra packages for the latter). Then "tackle" the following questions:

-   Do you understand what each line means? Which lines are (particularly) confusing?
-   Check out the questions following each code block and try to answer them.

## Troubleshooting

1.  If it all looks very confusing right now and you cannot answer any of the questions, do not despair! We will go through everything step by step during the practical and you will soon become familiar with such code. Note that we do **not** expect you to be able to answer all these questions today!
2.  If you cannot execute the code at all, tell us as soon as you can, so we can try to fix this together. Everyone should be able to run the code on their computer - it's not sufficient to peak onto your neighbor's screen.
3.  A great tool to understand what a line of code does is to print the output line by line, using the `print()` command. Be aware that in long loops this can create quite a mess of output.

------------------------------------------------------------------------

## Simulating evolutionary rescue in discrete time

In this notebook, we are implementing simulations of evolutionary rescue by a single mutation in discrete time, as discussed and analyzed in the work by Orr & Unckless. The model is as follows: a haploid population starts with a given number of wild-type and mutant individuals. Because of a sudden environmental change, the wild type is unfit and declines in the new environment. The mutant is beneficial and can survive and expand in the new environment, but it starts at a low (or zero) frequency in the total population. In order to rescue the population, it has to outcompete the wild type. If the mutant is not present at time 0, it has to appear through mutation and then spread in the population before extinction occurs.

## List of variables and parameters

-   $N_a$ current size of the wild-type (a) population (changes over time)
-   $N_A$ current size of the mutant (A) population (changes over time)
-   `init_wm` initial population size of wild-type population (should be a positive integer)
-   `init_wm` initial population size of the mutant population (should be a non-negative integer)
-   `decay_rate` detrimental effect of the environment on the wild type (i.e., the wild-type population decays at this rate; should be \>0 to induce potential extinction)
-   `sel_coeff` selective advantage of the mutant over the wild type (should be \>decay_rate to allow for evolutionary rescue)
-   `mut_rate` mutation rate from the wild type to the mutant genotype (should be \>0 and small)
-   `t_max` number of generations after which to stop the simulation (should be an integer \>0; this is not a model parameter but a setting for the simulations). In the code presented here, the simulation stops automatically if the population goes extinct, or if it recovers and exceeds 1.5 times the original total population size, so `t_max` can be set to a large value without creating extremely long simulation times.

## Simulate one generation of population size change

We want to observe the numbers of wild-type and mutant individuals over time. As first step, we simulate one generation of the population. At time $t$, the population has $N_a$ wild-type individuals and $N_A$ mutant individuals. First, these reproduce. This is modeled by drawing a Poisson random variable according to the expected absolute number $N_a \cdot (1-r)$ of wild-type individuals, where $r$ is the decay rate of the wild type, and $N_A \cdot (1-r+s)$ of mutant individuals, where $s$ is the selective advantage of the mutant over the wild time. Then mutation from $a$ to $A$ occurs as a Poisson random variable according to the given mutation rate. This defines the composition of the population at time $t+1$.

## Our Question

How do 2 deleterious mutations of different strength , which can mutate to beneficial ones, influence the probabilty of rescue and how does their frequency in the population change?

## Definitions
- wd: weak deleterious
- wb: weak beneficial
- sd: strong deleterious
- sb: strong beneficial

```{r}
# Function for simulating one Generation in the Population
simulate_one_gen <- function(N_wd, N_wb, N_sd, N_sb, decay_rate_wm, decay_rate_sm, s_wm, s_sm, m_rate_wm, m_rate_sm) {
  
  # draw Offspring according to Poisson Distribution
  offsp_wd <- rpois(1, N_wd * (1-decay_rate_wm-s_wm))
  offsp_wb <- rpois(1, N_wb * (1-decay_rate_wm+s_wm))
  offsp_sd <- rpois(1, N_sd * (1-decay_rate_sm-s_sm))
  offsp_sb <- rpois(1, N_sb * (1-decay_rate_sm+s_sm))
  # draw changing Mutants according to Poisson Distribution
  mut_wd_to_wb <- rpois(1, offsp_wd * m_rate_wm)
  mut_sd_to_sb <- rpois(1, offsp_sd * m_rate_sm)
  mut_wb_to_wd <- rpois(1, offsp_wb * m_rate_wm)
  mut_sb_to_sd <- rpois(1, offsp_sb * m_rate_sm)
  
  # determine new Population Sizes of all four Mutants (2 versions of weak Mutations   and 2   versions of strong Mutations)
  N_wd_new <- max(offsp_wd - mut_wd_to_wb, 0)+(offsp_wd + mut_wb_to_wd)
  N_wb_new <- max(offsp_wb - mut_wb_to_wd, 0)+(offsp_wb + mut_wd_to_wb)
  N_sd_new <- max(offsp_sd - mut_sd_to_sb, 0)+(offsp_sd + mut_sb_to_sd)
  N_sb_new <- max(offsp_sb - mut_sb_to_sd, 0)+(offsp_sb + mut_sd_to_sb)
  
 return(c(N_wd_new, N_wb_new, N_sd_new, N_sb_new))
}
# Test the Function
print(simulate_one_gen(N_wd=50, N_wb=0, N_sd=50, N_sb=0, decay_rate_wm=0.1, decay_rate_sm=0.2, s_wm=0.2, s_sm=0.4, m_rate_wm=0.1, m_rate_sm=0.1))
```

### Questions

-   Why do we substract `mut_a_to_A` from `N_a_new` and not from `N_A_new`? a mutates to A, therefore these individuals are lost in a and gained in A
-   In the life cycle as defined above, do individuals reproduce first and then mutate, or vice versa? reproduce and then mutate
-   Running the test output repeatedly, why do you almost always get mutant individuals in the next generation, even though you start without any mutants? Mutation rate is 1 % and its assumed that all 100 individuals reproduce.
-   What does the function `c` do? It puts them together in an object to get returned.

## Simulate a population trajectory

We can now track the dynamics of the population from one generation to another, but this is not enough. We want to follow the population from time 0 until it either is rescued or goes extinct. This is what the following function does.

```{r}
# Simulation runs until the 100th Generation
max_gen=1000
simulate_pop <- function(init_wd, init_wb,  init_sd, init_sb, decay_rate_wm, decay_rate_sm, s_wm, s_sm, m_rate_wm, m_rate_sm, max_gen) {
  # Create the Vector in which to save the Results
  pop_vector <- c(init_wd,init_wb, init_sd, init_sb)
  # initiate the variables
  pop_new <- c(init_wd, init_wb, init_sd, init_sb)
  
  # run the Simulation until Generation max_gen
  for (i in 1:(max_gen + 1)) {
    # redefine the current Population one Generation later
    pop_new <- simulate_one_gen(pop_new[1],pop_new[2],pop_new[3],pop_new[4], decay_rate_wm, decay_rate_sm, s_wm, s_sm, m_rate_wm, m_rate_sm)
    # add the new Population Sizes to the Output Vector
    pop_vector <- rbind(pop_vector,pop_new)
    # Condition to stop the Simulation before max_gen: either the Population exceeds 1.5         Times the Original Population Size, or it goes extinct
    if ((sum(pop_new) >= 10 * sum(c(init_wd, init_wb, init_sd, init_sb))) || sum(pop_new) == 0)
      break

}
  
  # Define the Row and Column Names of the Output Vector
  rownames(pop_vector) <- (0:max_gen)[1:length(pop_vector[,1])] # Note that the Vector has to be cut if the Simulation stopped early
  colnames(pop_vector) <- c("wd","wb", "sd", "sb")
  # return the Result
  return(pop_vector)	
}


# Test the Function and plot the Result
# set max_gen to a Value to be used througout
max_gen <- 1000
# create your Simulation Data
output <- simulate_pop(init_wd=100, init_wb=0,  init_sd=100, init_sb=0, decay_rate_wm=0.2, decay_rate_sm=0.4, s_wm=0.1, s_sm=0.4, m_rate_wm=0.01, m_rate_sm=0.1,max_gen)
# show the last few Lines of the Data Table
print(tail(output))
# plot the Output - note that the Plot Range and the x Axis have to be adjusted based on the Length of the Table
# determine x axis Range 
x_range <- 0:(length(output[,1])-1)
# this plots the total Population Size
plot(x_range,output[,1]+output[,2]+output[ ,3]+output[ ,4],type='l',ylim=c(0,max(output[,1]+output[,2]+output[ ,3]+output[ ,4])),xlab =  "Generation",ylab = "Population size")
# add Number of wd Individuals
lines(x_range,output[,1], col="blue")
# add Number of wb Individuals
lines(x_range,output[,2], col="navyblue")
# add Number of sd Individuals
lines(x_range,output[,3], col="red")
# add Number of sb Individuals
lines(x_range,output[,4], col="darkred")
# add Legend
legend("topleft",
       legend=c("Weak deleterious","Weak beneficial","Strong deleterious","Strong beneficial"),
       col=c("blue","navyblue","red","darkred"), lty=1, lwd=1.5)
```

### Questions

-   What is the difference between `pop_vector` and `pop_new`? In pop_vector the results are saved and in pop_new is the function applied that simulates one generation. So in pop_new the function is run again and again, therefore the results are changed and overwritten. But they are saved in pop_vector, therefore pop_vector later contains a list of different results
-   What is the condition that breaks the loop good for? What are its disadvantages? So that the loop doesn't run to max. Also you could probably detect errors better and save some time. It could fail to show all results, for example if the population gets rescued but the populations size doesnt exceed 1.5 \* population size before. Could get tables of different lengths =\> more difficult to analyse it together.
-   What does the function `tail` do? It shows the last results of the function. To check that something happened and it does what you expect it to do.

## A small analysis

As you can see if you simulate and plot the output several times, there is a lot of stochasticity/variation in the outcome. Therefore, to get interpretable results, we have to run the simulation many times and summarize the outcome depending on the question of our study. Here I am showing an example in which we record the time and the value of the lowest population size (or the time of extinction, if the population size eventually goes to 0). Usually, we want to study this for a range of model parameters; here I am varying the decay rate and the selection coefficient. Other (maybe more interesting) features to study would be the following: =\>simulating varying parameters with replicates per one value

-   rescue probability dependent on initial frequency and selection coefficient
-   time at which the rescue mutation occurs first if it rescues, vs. if it doesn't rescue, dependent on initial population size and mutation rate
-   time until the population has recovered in case it becomes rescued, dependent on mutation rate and selection coefficient
-   ...

```{r}
# set some Parameters to fixed Values
init_wd <- 50
init_wb <- 0
init_sd <- 50
init_sb <- 0
m_rate_wm <- 0.001
m_rate_sm <- 0.001
decay_rate_wm <- 0.1
decay_rate_sm <- 0.3
s_wm <- 0.2
s_sm <- 0.4
max_gen <- 1000
# determine how often to run the Simulation for each set of Parameters
no_replicates <- 100

# set Parameters to vary
s_values_wm <- c(0.2,0.3,0.4)
s_values_sm <- c(0.3,0.4,0.5)
m_values_wm <- c(0.01,0.05,0.1)
m_values_sm <- c(0.1,0.15,0.2)

# initialize Data Table - where to collect the Results
data_table <- c()
# run the Simulation across all chosen Parameters
# loop over decay rates
for(mval_wm in m_values_wm){
  for(mval_sm in m_values_sm){
  # loop over selection coefficients
    for(sval_wm in s_values_wm){
      for(sval_sm in s_values_sm){
    # different way of running many Simulations: make replicates using "repeat" Function with a        counter i
    # reset Counter
      i<-1
      repeat {
      # increase Counter by one
        i<-i+1
      # run the Simulation once
        one_run <- simulate_pop(init_wd, init_wb, init_sd, init_sb,
                        decay_rate_wm, decay_rate_sm,
                        sval_wm, sval_sm, mval_wm, mval_sm, max_gen)
      # determine total Population Sizes
        total_size <- one_run[,1]+one_run[,2]+one_run[,3]+one_run[,4]
      # determine minimum Population Size
        min_size <- min(total_size)
      # determine (first) Generation at which this population size occurred
        min_gen <- as.numeric(which(total_size==min_size)[1])
      # enter the Data into the Table
        data_table <- rbind(data_table,c(mval_wm,mval_sm,sval_wm,sval_sm,min_gen,min_size)) 
        # note that we add the varying Parameters (selection strength and mutation rate for          each weak and strong Mutation) to the Table too
      # stop the repeated Computation after no_replicates times
        if(i>no_replicates) break
    }
  }
  }
  }
}
# define Column Names of Data Table
colnames(data_table) <- c("m_rate_wm","m_rate_sm","s_wm","s_sm","min_gen","no_min")
# show the first and last Lines of the Output
print(head(data_table))
print(tail(data_table))
```

### Questions

-   Why does this code block run much more slowly than the others? Because its 3 loops intertwined together: First the for then again a for function in the previous for function and then the repeat function (and repeats)
-   Why is it useful to define the parameters at the beginning (e.g. `init_a=100`) and use their names rather than the numbers throughout the rest of the code? So that you can change the numbers easily, otherwise you would have to change the number everytime it appears in the code.
-   What does the function `which` do? It only chooses variables (its position) that fulfil a certain condition.

## Next steps

So, where do you go from here? Of course you won't just stare at these tables (at least not forever). Visualization would be a good step to see how the minimum population size varies with the decay rate and the selection coefficient. And you might want to include only the cases in which rescue happens, and also extract the rescue probability in parallel. However, the idea is that you pick a question that you find (more) interesting and implement modified versions of this code to address your question of choice.
